---
output:
  html_document: default
  word_document: default
  pdf_document: default
---
#GRAPH THEORY APPLICATIONS TO COMPREHEND EPIDEMICS SPREAD OF A DISEASE
#Disease outbreaks that are infectious form a repeating menace both to animals and humans.
#It has devastating effect on all aspects of biodiversity, health and economy.
#The best strategy to contain it is fast determination of cases forming clusters and mitigation follow up of the strategy to contain the epidemic spread.                                                                          
#1.Spatial      - spatial kernel
#2.Temporal     - time elapsed
#3.Genetic WGS  - rate of mutation

```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7, 
  fig.height = 5,
  out.width = "80%"
)

```


```{r}
load("D:/Naran/Sem 5/SNDA/Package/rabies_car_2003.RData")

rabies_car_2003
current_indices <- row.names(rabies_car_2003$dna)
indices_to_remove <- rabies_car_2003$linelist$index[is.na(rabies_car_2003$linelist$latitude) & is.na(rabies_car_2003$linelist$longitude)]

rabies_car_2003$dna <- rabies_car_2003$dna[-which(current_indices %in% indices_to_remove), , drop = FALSE]

valid_rows <- !is.na(rabies_car_2003$linelist$latitude) & !is.na(rabies_car_2003$linelist$longitude)

rabies_car_2003$linelist <- rabies_car_2003$linelist[valid_rows, ]

print(rabies_car_2003$linelist)

```


```{r}
rabies_car_2003
```

```{r date}

library(incidence) 
## weekly
plot(incidence(rabies_car_2003[["linelist"]][["date"]],7))

## monthly
plot(incidence(rabies_car_2003[["linelist"]][["date"]],365/12))

```
```{r locations}
#locations <- t(sapply(1:sim_rabies$n, function(i) sim_rabies$cases[[i]]$location))
#plot(locations, xlab = "", ylab = "")
plot(rabies_car_2003[["linelist"]][["longitude"]], rabies_car_2003[["linelist"]][["latitude"]],
     xlab = "Longitude",  # Set x-axis label
     ylab = "Latitude",   # Set y-axis label
     main = "Scatter Plot of Locations",
     pch = 19,            # Point type
     col = "blue")        # Color of the points
```
```{r dna}
dna<-rabies_car_2003[["dna"]]
class(dna)
length(dna)

dna_sequence <- as.character(dna)
#print(dna_sequence[1:5,1:5])
```
```{r Distance_dates}
head(rabies_car_2003[["linelist"]][["date"]], 5)
D_dates <- dist(rabies_car_2003[["linelist"]][["date"]])
#D_dates
class(D_dates)
```

```{r Distances_locations}
longitudes <- rabies_car_2003[["linelist"]][["longitude"]]
latitudes <- rabies_car_2003[["linelist"]][["latitude"]]
locations <- cbind(longitudes, latitudes)

library(fields)
#print(latitudes)
#print(longitudes)
D_geo <- rdist(locations)
#D_geo
```
```{r Distances_dna}
library(ape)
D_hamming <- dist.dna(dna, model = "raw")

# Print the resulting distance matrix
#print(D_hamming)

D_hamming_raw <- dist.dna(dna, model = "N", pairwise.deletion = FALSE)

# Print the raw Hamming distance matrix (actual number of differing positions)
print(D_hamming_raw)

D_dna<-D_hamming_raw
print(D_dna)
```

```{r}
library(ape)
tree <- ladderize(root(nj(D_dna), 1))
plot(tree)
axisPhylo()
mtext(side = 1, text = "Number of mutations", line = 2.5)
```
```{r distances, fig.height = 7} 
library("vimes")
D_all <- vimes_data(dates = D_dates, geo = D_geo, dna = D_dna)
plot(D_all, nclass = 60)
``` 
# Defining cutoff distances above which cases are considered not linked by transmission 

## Distributions of expected distances between cases for rabies

The distribution of expected distances between cases in the temporal, spatial, and
genetic spaces.
```{r}
gamma_mean <- mean(D_dates)
gamma_std <- sd(D_dates)

gamma_shape <- gamma_mean^2 / (gamma_std^2)
gamma_scale <- gamma_std^2 / gamma_mean

print(gamma_shape)
print(gamma_scale)
```


```{r}
#D_geo
rayleigh_mean <- mean(D_geo)
rayleigh_scale <- rayleigh_mean / sqrt(acos(-1)/2)

rayleigh_mean
rayleigh_scale
```
```{r}
n_sites <- ncol(rabies_car_2003$dna)
mu_year_per_site <- 5.9e-4
#mu_year_per_site <- observed_mutations / (n_sites * years)

mu_day_whole <- (mu_year_per_site * n_sites / 365)

gamma_mean_dna <- mean(D_dna)
gamma_std_dna <- sd(D_dna)

gamma_shape_dna <- gamma_mean_dna^2 / (gamma_std_dna^2)
gamma_scale_dna <- gamma_std_dna^2 / gamma_mean_dna

print(gamma_shape_dna)
print(gamma_scale_dna)

```

```{r plot_f, fig.height = 7} 
## distance functions for each of the 3 types
f_temporal <- fpaircase(type = "temporal", gamma_shape = gamma_shape,
                        gamma_scale = gamma_scale)

f_spatial <- fpaircase(type = "spatial", sd_spatial = rayleigh_scale)

f_genetic <- fpaircase(type = "genetic", poisson_rate = mu_day_whole,
                       gamma_shape = gamma_shape_dna, 
                       gamma_scale = gamma_scale_dna)

## Plotting these
par(mfrow=c(3,1))
plot(f_temporal, xlim = c(0,365))
plot(f_spatial, xlim = c(0,5))
plot(f_genetic, xlim = c(0,9))
```

```{r}
## reporting rate
pi <- 0.2

## quantiles
q <- c(.50, .75, .90, .95, .95^(1/3), .99, .999, .9995, .9999)

## colours used to plot these
cols <- rainbow(length(q))

## our main results are with the cutoff corresponding to the 95% quantile
cutoff_choice <- q[4] 
```

```{r}
plot(D_dates,f_temporal(D_dates), main = "Temporal Function with Quantile Cutoffs", 
     xlab = "Temporal Distance (days)", ylab = "Density")

cutoffs <- qgamma(q, shape = gamma_shape, scale = gamma_scale)
y_vals = f_temporal(D_dates)
# Add vertical lines and labels for each quantile
for (i in seq_along(q)) {
  abline(v = cutoffs[i], col = cols[i], lwd = 2, lty = 2)
  
  # Adjust y position for visibility
  label_y <- max(y_vals) * 0.85  # Adjust as needed
  
  # Print cutoff and y position for debugging
  print(paste("Cutoff:", cutoffs[i],  "Quantile : ",q[i]))
  
  # Add the text label
  text(cutoffs[i], label_y, labels = paste("q =", q[i]), pos = 3, col = cols[i], cex = 0.8)
}
```
```{r}
# Calculate the cutoff distances based on D_geo
cutoffs <- quantile(D_geo, probs = q)

# Create a sequence of x values for plotting
x_vals <- seq(0, max(D_geo) + 1, length.out = 100)
y_vals <- f_spatial(x_vals)
# Plot the spatial function
plot(x_vals, f_spatial(x_vals), type = "l", col = "blue", lwd = 2,
     xlim = c(0, max(D_geo) + 1), ylim = c(0, max(f_spatial(x_vals))),
     xlab = "Distance", ylab = "Density", main = "Spatial Function with Quantiles")

# Add vertical lines for each quantile cutoff
for (i in seq_along(cutoffs)) {
    abline(v = cutoffs[i], col = rainbow(length(q))[i], lwd = 2, lty = 2)  # Dashed lines
  
    # Print cutoff and y position for debugging
    print(paste("Cutoff:", cutoffs[i],  "Quantile : ",q[i]))
    # Add labels for quantiles
    #text(cutoffs[i], max(f_spatial(x_vals)) * 0.9, labels = paste("q =", round(q[i], 3)), pos = 3, col = rainbow(length(q))[i])
}

# Optionally add a legend
legend("topright", legend = paste("q =", round(q, 3)), 
       col = rainbow(length(q)), lwd = 2, lty = 2)

```

```{r}
# Plot the genetic function
plot(D_dna,f_genetic(D_dna), col = "blue", lwd = 2,
     xlab = "Genetic Distance", ylab = "Density",
     main = "Genetic Distance Function with Quantiles")

cutoffs <- quantile(D_dna, probs = q)

# Add vertical lines for each quantile cutoff
for (i in seq_along(cutoffs)) {
  abline(v = cutoffs[i], col = cols[i], lwd = 2, lty = 2)  # Dashed lines for quantiles
  
  print(paste("Cutoff:", cutoffs[i],  "Quantile : ",q[i]))
  
  # Add a label at the top of the plot for each cutoff
  #text(cutoffs[i], max(y_vals), labels = paste("q =", q[i]), pos = 3, col = cols[i])
}
legend("center", legend = paste("q =", round(q, 3)), 
       col = rainbow(length(q)), lwd = 2, lty = 2)
```

```{r results, cache = TRUE, fig.width = 9, fig.height = 9, out.width = "100%"}
### function used to get results for a certain cutoff and reporting rate

get_res <- function(D_all, q, pi, f_temporal, f_spatial, f_genetic,
                    type = c("all", "temporal","spatial", "genetic")) {
  
  type <- match.arg(type)
  
  ## get the cutoffs
  cuts <- c(temporal = get_quantiles(f_temporal, q, pi = pi), 
            spatial = get_quantiles(f_spatial, q, pi = pi),
            genetic = get_quantiles(f_genetic, q, pi = pi))
  
  if (type == "all") {
    ## use vimes
    out <- vimes(D_all, cutoff = cuts,
                 graph.opt = vimes.graph.opt(col.pal = funky))
  } else if (type == "temporal") {
    out <- vimes(vimes_data(dates = D_all$dates), cutoff = cuts["temporal"],
                 graph.opt = vimes.graph.opt(col.pal = funky))
  } else if (type == "spatial") {
    out <- vimes(vimes_data(geo = D_all$geo), cutoff = cuts["spatial"],
                 graph.opt = vimes.graph.opt(col.pal = funky))
  } else if (type == "genetic") {
    out <- vimes(vimes_data(dna = D_all$dna), cutoff = cuts["genetic"],
                 graph.opt = vimes.graph.opt(col.pal = funky))
  }
    
  return(out)
  
}

### Use the function above to generate results for several combinations of p and pi in combi (below)
combi <- expand.grid(p = q,
                     pi = pi)
combi

quantile_pretty <- signif(combi$p*100, 4)
quantile_pretty <- paste0(quantile_pretty, "%")

res <- vector(9L, mode = "list")

for (i in 1:nrow(combi)) {
    res[[i]] <- get_res(D_all, combi[i, 1],
                        combi[i, 2], f_temporal,
                        f_spatial, f_genetic)
}

### Visualizing the output
par(mfrow = c(3, 3), mar=c(1,1,3,1))

for (i in 1:length(res)) {
  plot(res[[i]]$graph, vertex.label = "",
       main = paste("cutoff:", quantile_pretty[i]))
}

```

# Estimate the underlying reproduction number and number of imported cases
```{r eval = FALSE} 
library(devtools)
install_github("reconhub/branchr")
```

```{r} 
library(branchr)

#Cluster size and reporting rates are taken as inputs for this function which computes the Reproduction number - R
compute_R  <- function(cl_size, rho) {
    profile <- profile_likelihood(y_obs = cl_size, 
                                  rho = rho, 0.01, 20)      #profile of how likely each value of R in the range 0.01 to 20 is
    R_estimate <- theta_max_likelihood(profile$theta,
                                       profile$Likelihood,  
                                       0.95)                #maximum likelihood estimate and a 95% confidence interval for R using the                                                                    #likelihood profile
    R <- c(central = R_estimate$theta_max_likelihood, 
           low = R_estimate$lower_theta, 
           up = R_estimate$upper_theta)                     #estimated central value of R and its lower and upper confidence bounds
    
    import <- import(y_obs = cl_size, 
                     rho = rho, 
                     profile, 1e3, 1e3, 0.95)               #estimated number of unobserved cases (importations) 
                                                            #1000 (first instance): number of simulations to estimate the unobserved cases
                                                            #1e3 (second instance): number of iterations for importation estimation process
                                                            #0.95 ( 95% )         : confidence level for the resulting estimates
    unobs <- c(central = import$theta_max_likelihood,
               low = import$lower_theta,
               up = import$upper_theta)                     #central, lower, and upper estimates of the unobserved cases
    
    return(list(R, unobs))
}

clust_size <- lapply(res, function(i) i$clusters$size)
rho <- combi[, 2]

R_estimate_and_imports <- lapply(1:length(clust_size),
                                 function(i) compute_R(clust_size[[i]], rho[i]))

R_estimates <- sapply(1:length(R_estimate_and_imports), 
                      function(i) R_estimate_and_imports[[i]][[1]])

N_unobs_estimates <- sapply(1:length(R_estimate_and_imports), 
                            function(i) R_estimate_and_imports[[i]][[2]])

N_tot_estimates <- N_unobs_estimates + 
  matrix(rep(lengths(clust_size), 3), 
         nrow = 3, byrow = TRUE)

n_days <- diff(range(rabies_car_2003[["linelist"]][["date"]]))

rate_import_tot_days <- N_tot_estimates / as.numeric(n_days)
rate_import_tot_year <- rate_import_tot_days*365


### plotting results ###

par(mfrow = c(2, 1))

## R estimates
plot(R_estimates["central",], ylim = c(0, max(c(1.2, max(R_estimates)))),
     pch = 19, xlab = "Cutoff used for pruning", ylab = "Estimated R", axes = FALSE)
axis(side = 1, at = 1:length(R_estimate_and_imports),
     labels = quantile_pretty, cex = 0.75)
axis(side = 2)
for (i in 1:length(R_estimate_and_imports)) {
  segments(i, R_estimates["low",i],i, R_estimates["up",i])
}

## adding the reproduction number used to generate the simulation
abline(h = 0.92, col = "red", lty = 2)

## importation rate estimates
plot(rate_import_tot_year["central",], ylim = c(0, max(rate_import_tot_year)),
     pch = 19, xlab = "Cutoff used for pruning",
     ylab = "Estimated rate of importation (per year)", axes = FALSE)
axis(side = 1, at = 1:length(R_estimate_and_imports),
     labels = quantile_pretty, cex = 0.75)
axis(side = 2)
for (i in 1:length(R_estimate_and_imports)) {
  segments(i, rate_import_tot_year["low",i],i, rate_import_tot_year["up",i])
}

# adding the importation rate used to generate the simulation
abline(h = 7, col = "red", lty = 2)

```
